---
title: "Analysis of United States Data Scientists in the 2018 Stack Overflow Survey"
author: "John Kane"
date: '2018-06-03'
slug: analysis-of-united-states-data-scientists-in-the-2018-stack-overflow-survey
tags:
- career
- stackoverflow
categories:
- career
- stackoverflow
---


```{r}
library(tidyverse)
load("/home/john/johnckane.com/drafts/stackoverflow-2018-survey/developer_survey_2018/us_ds.Rda")
#colnames(us_ds)
```

This is an analysis of the [2018 Stack Overflow Survey](link). 

This is limited to those identifying as a "Data scientist or machine learning specialist," living in the United States, and employed full-time.

I am a data scientist and participated in the survey. I hoped to see how I compared to my peers with respect to questions of interest to me that relate to career and lifestyle.

As a companion to this post I've created Shiny Application that allows users to input their own responses (no data is saved) to see how they compare to
other data scientists in the United States.


There is a lot more information in the SO Survey and I encourage you to check it out. This analysis reflects the questions that were of most interest to me this week.


Jump to:

Background
Toolkit
Current Job and Career (including salary)
Lifestyle
Imposter Syndrome

```{r}
plot_func <- function(data,variable,other,title, offset = 0.05){

  func_df <- data
 
  if(other == TRUE){
    lt1_names <- names(which(prop.table(table(func_df[[variable]])) < .01)) 
    lt1_names_index <- which(func_df[[variable]] %in% lt1_names)
    func_df[lt1_names_index,variable] <- "Other"
  }
  
  
  na_index <- which(is.na(func_df[[variable]]))
  func_df[na_index,variable] <- "NA"
  var_levels <- names(table(func_df[[variable]]))[order(table(func_df[[variable]]))]

  func_df <- func_df %>%
    mutate(factor_var = factor(eval(parse(text=variable)), levels = var_levels)) %>%
    group_by(factor_var) %>%
    summarise(total_level = n()) %>%
    mutate(dummy = 1) %>%
    inner_join(func_df %>%
                 summarise(total = n()) %>%
                 mutate(dummy = 1),
               by = "dummy") %>%
    mutate(pct = total_level/total)

  ggplot(data = func_df,
         aes(x = factor_var,
             y = pct)) +
    geom_bar(stat = "identity",fill = "white",color = "darkgreen") +
    coord_flip() +
    theme_minimal() +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    scale_x_discrete("") +
    ggtitle(title) +
    geom_text(aes(x = factor_var, y = pct + offset,
                  label = paste0(round(100*round(pct,2)),"%"))) +
    scale_y_continuous("",
                        breaks = NULL,
                        labels = NULL)
}
```


## Background

```{r,warning=FALSE}
plot_func(data = us_ds, variable = "Age", other = FALSE, title = "Age", offset = 0.02)
```

```{r}
plot_func(data = us_ds, variable = "Gender", other = TRUE, title = "Gender")
```

```{r}
plot_func(data = us_ds, variable = "EducationParents", other = TRUE, title = "Parent's Education")
```


```{r}
plot_func(us_ds,"FormalEducation",other = TRUE, title = "Formal Education")
```


```{r}
plot_func(us_ds,"UndergradMajor",TRUE,title = "Undergraduate Major", offset = 0.03)
```


```{r}
plot_func(us_ds,"YearsCodingProf",other = TRUE,title ="Years Coding Profesionally", offset = 0.02)
```


## Toolkit

```{r}
plot_func2 <- function(data,variable,other,sample_size,title,offset = 0.05){

  func_df <- data
 
  if(other == TRUE){
    lt1_names <- names(which(table(func_df[[variable]])/sample_size < .01)) 
    lt1_names_index <- which(func_df[[variable]] %in% lt1_names)
    func_df[lt1_names_index,variable] <- "Other"
  }
  
  
  na_index <- which(is.na(func_df[[variable]]))
  func_df[na_index,variable] <- "NA"
  var_levels <- names(table(func_df[[variable]]))[order(table(func_df[[variable]]))]

  func_df <- func_df %>%
    mutate(factor_var = factor(eval(parse(text=variable)), levels = var_levels)) %>%
    group_by(factor_var) %>%
    summarise(total_level = n()) %>%
    mutate(pct = total_level/sample_size)

  ggplot(data = func_df,
         aes(x = factor_var,
             y = pct)) +
    geom_bar(stat = "identity",fill = "white",color = "darkgreen") +
    coord_flip() +
    theme_minimal() +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    scale_x_discrete("") +
    ggtitle(title) +
    geom_text(aes(x = factor_var, y = pct + offset,
                  label = paste0(round(100*round(pct,2)),"%"))) +
    scale_y_continuous("",
                        breaks = NULL,
                        labels = NULL)
}

```

# What else do data scientists do?

```{r}
dt <- str_split(us_ds$DevType,";") %>% unlist() %>% unname()
dt <- dt[!(is.na(dt) | dt=="Data scientist or machine learning specialist")] 
dt_df <- data.frame(devtype = dt, stringsAsFactors = FALSE)
plot_func2(dt_df, "devtype", other = TRUE, sample_size = sum(!is.na(us_ds$DevType)), title = "Developer Type (besides Data Scientist)")
```




       
       
       
       
       
       
```{r}
lww <- str_split(us_ds$LanguageWorkedWith,";")
lww <- lww[!is.na(lww)] %>% unlist() %>% unname()
lww_df <- data.frame(Language = lww, stringsAsFactors = FALSE)
plot_func2(lww_df, "Language", other = TRUE, sample_size = sum(!is.na(us_ds$LanguageWorkedWith)), title = "Coding Languages")
```
       
       
```{r}
et <- str_split(us_ds$EducationTypes,";")
et <- et[!is.na(et)] %>% unlist() %>% unname()
et_df <- data.frame(EdType = et, stringsAsFactors = FALSE)
plot_func2(et_df,"EdType",other = TRUE, sample_size = sum(!is.na(us_ds$EducationTypes)), title = "Education Types", offset = 0.1)
```
       
       
```{r}
dbww <- str_split(us_ds$DatabaseWorkedWith,";")
dbww <- dbww[!is.na(dbww)] %>% unlist() %>% unname()
dbww_df <- data.frame(Database = dbww, stringsAsFactors = FALSE)
plot_func2(dbww_df,"Database",other=TRUE,sample_size = sum(!is.na(us_ds$DatabaseWorkedWith)), title = "Databases")
```
       
```{r}
pww <- str_split(us_ds$PlatformWorkedWith,";")
pww <- pww[!is.na(pww)] %>% unlist() %>% unname()
pww_df <- data.frame(Platform = pww, stringsAsFactors = FALSE)
plot_func2(pww_df, "Platform",other=TRUE,sample_size = sum(!is.na(us_ds$PlatformWorkedWith)), title = "Platforms", offset = 0.03)
```
       
       
```{r}
fww <- str_split(us_ds$FrameworkWorkedWith,";")
fww <- fww[!is.na(fww)] %>% unlist() %>% unname()
fww_df <- data.frame(Framework = fww, stringsAsFactors = FALSE)
plot_func2(fww_df, "Framework",other=TRUE,sample_size = sum(!is.na(us_ds$FrameworkWorkedWith)), title = "Frameworks", offset = 0.02)
```

```{r}
ide <- str_split(us_ds$IDE,";")
ide <- ide[!is.na(ide)] %>% unlist() %>% unname()
ide_df <- data.frame(IDE = ide, stringsAsFactors = FALSE)
plot_func2(ide_df, "IDE", other = TRUE, sample_size = sum(!is.na(us_ds$IDE)), title = "IDE", offset = 0.02)
```
       
```{r}
os <- str_split(us_ds$OperatingSystem,";")
os <- os[!is.na(os)] %>% unlist() %>% unname()
os_df <- data.frame(OS = os, stringsAsFactors = FALSE)
plot_func2(os_df, "OS", other = TRUE, sample_size = sum(!is.na(us_ds$OperatingSystem)), title = "OS", offset = 0.02)
```

```{r}
nm <- str_split(us_ds$NumberMonitors, ";")
nm <- nm[!is.na(nm)] %>% unlist() %>% unname()
nm_df <- data.frame(NM = nm, stringsAsFactors = FALSE)
plot_func2(nm_df, "NM", other = TRUE, sample_size = sum(!is.na(us_ds$NumberMonitors)), title = "Number of Monitors")
```

```{r}
vc <- str_split(us_ds$VersionControl, ";")
vc <- vc[!is.na(vc)] %>% unlist() %>% unname()
vc_df <- data.frame(VC = vc, stringsAsFactors = FALSE)
plot_func2(vc_df, "VC", other = TRUE, sample_size = sum(!is.na(us_ds$VersionControl)), title = "Version Control")

```



## Current Job

```{r}
plot_func(us_ds,"CompanySize",TRUE,title = "CompanySize")
```

```{r}
plot_func(us_ds, "CompanySize", TRUE, title = "Company Size", offset = 0.02)
```

```{r}
plot_func(us_ds, "JobSatisfaction",TRUE,"Job Satisfaction", offset = 0.02)
```

```{r}
plot_func(us_ds,"JobSearchStatus",TRUE,title = "Job Search Status")
```

```{r}
plot_func(us_ds,"LastNewJob",TRUE,"Last New Job")
```

```{r}
plot_func(us_ds,"CareerSatisfaction",TRUE,title = "Career Satisfaction", offset = 0.02)
```


```{r}
ggplot(data = us_ds %>% filter(!is.na(us_ds$ConvertedSalary)),
       aes(x = ConvertedSalary)) +
  geom_histogram(binwidth = 5000, color = "darkgreen") +
  ggtitle("Salary", subtitle = paste0(as.character(sum(!is.na(us_ds$ConvertedSalary))), " Respondents")) +
  theme_minimal() +
  NULL
```

```{r}
ggplot(data = us_ds %>% filter(!is.na(ConvertedSalary),
                               ConvertedSalary < 500000),
       aes(x = ConvertedSalary)) +
  geom_histogram(binwidth = 5000, color = "darkgreen") +
  ggtitle("Salary", subtitle = paste0(as.character(sum(!is.na(us_ds$ConvertedSalary))), " Respondents")) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(from = 0,to=500000,by=25000),
                     labels = paste0("$",seq(from = 0,to=500,by=25),"k")) +
  NULL
```

```{r}
summary(us_ds$ConvertedSalary)
```


## Lifestyle
```{r}
plot_func(us_ds,"WakeTime",TRUE,title="On days when you work, what time do you typically wake up?")
```

### On a typical day...

```{r}
plot_func(us_ds,"HoursComputer",TRUE,title = "How much time do you spend on a desktop or laptop computer?")
```

```{r}
plot_func(us_ds,"HoursOutside",TRUE,title="How much time do you spend outside?")
```

### In a typical week...

```{r}
plot_func(us_ds,"Exercise",TRUE, title = "How many times do you exercise?")
```


```{r}
plot_func(us_ds,"SkipMeals",TRUE,title="How many times do you skip a meal in order to be more productive?")
```

## Community

Three questions in the survey I think can be interpreted as measuring the phenomenon known as [Imposter Syndrome](https://en.wikipedia.org/wiki/Impostor_syndrome). 

```{r}
plot_func
plot_func(us_ds,"AgreeDisagree1",other = FALSE, title =  "I feel a sense of kinship or connection to other developers")
```


> I think of myself as competing with my peers

```{r}
plot_func(us_ds,"AgreeDisagree2",other = FALSE, title =  "I think of myself as competing with my peers")
```

> I'm not as good at programming as most of my peers

```{r}
plot_func(us_ds,"AgreeDisagree2",other = FALSE, title =  "I'm not as good at programming as most of my peers")
```



```{r}
(100-16)/5
```




