---
title: NCAA Men's Tournament First Round Upsets
author: John Kane
date: '2018-04-08'
slug: ncaa-first-round-upsets
categories:
  - R
  - basketball
  - March Madness
tags:
  - R
  - basketball
  - March Madness
---

Unless you follow mid-major conferences closely, when it comes time to picking first round upsets in the NCAA Tournament you're likely either picking teams at random, or going off the analysis of the "experts" on TV, the internet, or podcasts. I won't address those who pick their upsets and Cinderella teams at random here, instead this post is about those who combine whatever they know about the field in a given year with what they see, read, and hear in the media. If you consume a lot this media you may be tempted to pick a lot of upsets, however by definition they should be rare.

Thanks to a dataset from a [Kaggle competition](https://www.kaggle.com/c/march-machine-learning-mania-2017/data) that provides tournament results from 1985-2016, I've put together an analysis that aims to offer guidance on how many upsets you could expect in the first round, and from which matchups they come from. I define an upset to be anytime a numerically higher seeds beats a numerically lower seed (e.g. a 10 seed beats a 7 seed). 



### Data Prep

```{r, include = FALSE}
data_path <- "/home/john/basketball/march_madness/2018/data/"
```


#### Load libraries and data
```{r, warning = FALSE, message = FALSE}
library(tidyverse)
library(magrittr)
library(knitr)
results <- read_csv(file = paste0(data_path,"TourneyCompactResults.csv"))
seeds <- read_csv(file = paste0(data_path,"TourneySeeds.csv"))
slots <- read_csv(file = paste0(data_path,"TourneySlots.csv"))
seasons <- read_csv(file = paste0(data_path,"Seasons.csv"))
colnames(results) <- tolower(colnames(results))
colnames(seeds) <- tolower(colnames(seeds))
colnames(slots) <- tolower(colnames(slots))
colnames(seasons) <- tolower(colnames(seasons))
```

#### Light data cleaning
```{r}
# convert colnames to lowercase
colnames(results) <- tolower(colnames(results))
colnames(seeds) <- tolower(colnames(seeds))
colnames(slots) <- tolower(colnames(slots))
colnames(seasons) <- tolower(colnames(seasons))

# parse the seed variable to get the region and numeric seed
seeds %<>%
  mutate(region = str_sub(seed,1,1),
         seed = as.numeric(str_sub(seed,2,3)))
```

#### Restructure the data
```{r}
results2 <-
  results %>%
  select(-daynum,-wscore,-lscore,-numot) %>% # drop uneeded variables
  inner_join(.,                              # append winning team seeds and regions
             seeds %>% rename(wseed = seed,wregion = region), 
             by = c("season","wteam" = "team")) %>%
  inner_join(.,                             # append losing team seeds and regions
             seeds %>% rename(lseed = seed,lregion = region),
             by = c("season","lteam" = "team")) %>%
  mutate(matchup = paste0(pmin(wseed,lseed),"-",pmax(wseed,lseed)),# create new variables indicating the matchup 
         upset =  ifelse(wseed > lseed, 1,ifelse(wseed == lseed, 0.5,0))) %>% # and if it was an upset
  select(season,matchup,upset,wregion,lregion)

head(results2)
```

### Historical Trends in Upsets

A point I want to address in the beginning is the way college basketball has changed during the period represented in the data. The best players in the country have gone from staying in school for three or four years to playing only a single season. Sometimes, it seems now, higher seeded teams that may not have a lot of NBA caliber talent, but have more experience and chemistry, could have a better chance of upsetting teams with rosters with less chemistry but more individual talent.

To consider this question further, we'll aggregate the data by season, then consider total number of upsets as well as upsets by matchup.

```{r}
total_upsets <- results2 %>%
    filter(matchup %in% c('1-16','2-15','3-14','4-13','5-12','6-11','7-10','8-9'), wregion==lregion) %>%
  group_by(season) %>%
  summarise(total_upsets = sum(upset))

by_matchup <- results2 %>%
  filter(matchup %in% c('1-16','2-15','3-14','4-13','5-12','6-11','7-10','8-9'), wregion==lregion) %>%
  group_by(season,matchup) %>%
  summarize(upsets = sum(upset))
```

The first plot considers total number of upsets in each tournament.

```{r}
ggplot(data = total_upsets,
       aes(x = season, y = total_upsets)) +
  geom_point(colour = "orange") +
  geom_line(colour = "orange") +
  geom_smooth(method = "loess") +
  scale_x_continuous("Season",minor_breaks = c(1985:2016)) +
  scale_y_continuous("Upsets", breaks = c(3:13)) +
  ggtitle("Number of Total First Round Upsets By Year", sub = "NCAA Men's Tournament, 1985-2016") +
  theme_minimal()
```
  
Some observations from this plot:

* For most of the tournament the best guess of the total number of upsets is somewhere between 7 and 9 (inclusive)
* 5 of the 8 times there have been at least 10 first round upsets have happened since 2009. 
* To that point, there is a slight trend upwards in recent years. However I won't jump to a conclusion that anything is necessarily different. But at the same time it could justify picking more upsets than the historical record would indicate.

The second plot considers the number of upsets by matchup over time.

```{r}
ggplot(data = by_matchup,
       aes(x = season, y = upsets)) +
  facet_wrap(~matchup,ncol = 4) +
  geom_line(colour = "orange") +
  geom_point(colour = "orange") +
  geom_smooth(method = 'loess') +
  scale_x_continuous("Season") +
  scale_y_continuous("Upsets") +
  ggtitle("Number of Upsets By Season and Matchup",sub = "NCAA Men's Tournament, 1985-2016") +
  theme_minimal()
```

Some observations from this plot:

* In the last few years there appears to be a trend for an increased likelihood of upsets in the 2-15 and 3-14 games going from zero to one upset.
* A trend for upsets in 5-12 games appears to have been a slow, ongoing phenomenon since the tournament began. Going from one in the 1980's to two today.
* A trend for an upset in 6-11 games appears to have begun back around the year 2000, going from one at the time to two today. 

Trends in the 2-15 and 3-14 games are very recent, whereas at least visually the 5-12 and 6-11 games appear to have more support. 

### Single Game Upset Probabilities

What is the chance of an upset in any one game? We can use the data to determine an empirical upset probability. 

```{r}
upset_prob <- 
results2 %>%
  filter(matchup %in% c('1-16','2-15','3-14','4-13','5-12','6-11','7-10','8-9'), wregion==lregion) %>% # Limit to 1st round games
  group_by(matchup) %>%
  summarise(upset_prob = sum(upset)/n())
upset_prob
```

One way to use this information to pick your bracket would be for each game roll a weighted die through the `sample()` function in R to determine if there was an upset or not. 

For example, to pick the winner of a 7-10 game (p = 0.39 for an upset), you would use:

```{r}
set.seed(123)
p <- 0.39
sample(c(1,0),1,prob = c(p,1-p))
```

Indicating you would not choose an upset for this game.

A second way to use the empirical upset probabilities would be to allow yourself a little discretion with who you choose to be upset. In the previous example you are the mercy of the die in each instance. You may end up picking a 3-14 upset, but not the one that seems most likely based on other datapoints (e.g. your own analysis, media takes) you receive. 

So instead what you can do is roll the die for all four matchups in a given tournament and then, if there are any upsets, you can choose which ones to fill in your bracket.

For example, you can pick the number of upsets in the four 7-10 matchups this way:

```{r}
set.seed(123)
p <- 0.39
sample(c(1,0),4,prob = c(p,1-p), replace = TRUE)
```

So in this case you would select, based on additional analysis, two of the four ten seeds to win. 

A valid critique of this approach is that you are still leaving a lot of things to chance, in the above example two upsets were selected, but the most likely outcome is actually for only one upset, albeit only slightly. You can confirm this with the following call to `dbinom()`:

```{r}
c(dbinom(1,4,0.39),dbinom(2,4,0.39))
```

So this approach may be leaving still too much to chance for many.

### Most Likely Number of Upsets

A less variable way to do this would be to choose how many upsets to select in given matchups based on historic tournment-wide results. For example, in how many of the 32 tournaments in this analysis has there been 1 upset in the 2-15 games? 

There are two constraints by this approach. The first is the total number of upsets, the second is from which matchups the upsets come from. 

##### Total First Round Upsets

What if we zoom out and consider how many total upsets there are in a given year. That would complement the number of upsets in each matchup and provide how many total upsets we may expect to see, and where we would expect to see them.

```{r}
ggplot(data = total_upsets,
       aes(x = total_upsets)) +
  geom_bar(fill = "orange") +
  scale_x_continuous("Total First Round Upsets",
                     breaks = c(3,4,5,6,7,8,9,10,11,12,13)) +
  scale_y_continuous("Percent") +
  ggtitle("Number of Total First Round Upsets", sub = "NCAA Men's Tournament, 1985-2016") +
  theme_minimal()
```

The obervations here are:
* There has always been at least 3 and never more than 13 total first round upsets.
* There have been between 7 and 10 (inclusive) upsets in 21/32 tournaments.
* You are more likely to have fewer than 7 upsets 7/32 than more than 10 (4/32). 

##### Number of upsets each year, by matchup

```{r}
ggplot(data = by_matchup,
       aes(x = upsets)) +
  facet_wrap(~matchup,ncol = 4) +
  geom_bar(fill = "orange") +
  scale_x_continuous("Upsets") +
  scale_y_continuous("Count") +
  ggtitle("Number of Upsets Each Year By Matchup",sub = "NCAA Men's Tournament, 1985-2016") +
    theme_minimal()

```

Some oberservations from this chart:
    * The most likely outcome every year is for there to be 0 upsets of 1 and 2 seeds, with at least one upset of each of teams seeded 3-7, with a tie of either 1 or 2 upsets of an 8 seed by a 9 seed.
    * There is more dispersion in the number of upsets as you move from the 1-16 to the 8-9 games.

### Suggested Approach to Picking First Round Upsets

Based on this analysis you could use the following heuristic for selecting how many upsets to pick and in which matchups to pick them. Of course which individual games to select requires some additional analysis.

So, in this case you would choose:

| Matchup | Recommended Upsets to Pick |
|---------|----------------------------|
| 1-16    | 0                          | 
| 2-15    | 0                          | 
| 3-14    | 0/1                        | 
| 4-13    | 1                          | 
| 5-12    | 1                          | 
| 6-11    | 1                          | 
| 7-10    | 1/2                        | 
| 8-9     | 1/2                        | 
|---------|----------------------------|
|Totals   | 5-8                        |

At minimum these recommendations will provide 5 upsets, but as we saw above there are usually between 7 and 10 upsets. The high end of this scale puts you into the range of most expected upsets, and recent trends may justify going upwards of 9 or 10. Based on historical results it seems best to source additional upsets from the 8-9, 7-10 or 6-11 games.

### Conclusion and Further Work

I think this analysis demonstrates that the likelihood of upsets 

In the course of doing this analysis I had some ideas for possible extensions:
* Extend this analysis into the Round of 32 and beyond
* compare chalk vs random brackets and points scored in traditional office-pool bracket scoring
* extend analysis into a Bayesian one, using appropriate prior and conjugate distributions 

### Postscript

The first round of the 2018 tournament has come and gone. Below is a table showing my recommendations for number of upsets along with how many upsets actually happened.

| Matchup | Recommended Upsets to Pick | Upsets in 2018 |
|---------|----------------------------|----------------|
| 1-16    | 0                          | 1              |
| 2-15    | 0                          | 0              |
| 3-14    | 0/1                        | 0              |
| 4-13    | 1                          | 2              |
| 5-12    | 1                          | 0              |
| 6-11    | 1                          | 1              |
| 7-10    | 1/2                        | 1              |
| 8-9     | 1/2                        | 3              |
|---------|----------------------------|----------------|
|Totals   | 5-8                        | 8              |

