---
title: An Equivalent to Convention for Batting Average
author: ''
date: '2020-01-14'
slug: an-equivalent-to-convention-for-batting-average
categories:
  - baseball
tags:
  - baseball
  - ops
---

The game of baseball has moved on statistically from batting average and RBIs. There are different metrics that are used to evaluate player skill and contribution. Moneyball came out 16 years ago, but I don't have a sense that the newer metrics.

If there were a single metric used to evaluate offensive player skill it would probably be WAR (Wins Above Replacement). The calculations that determine WAR vary by the source (Baseball Reference, Fan Graphs, ESPN, etc.) and are closed.

A different metric that can easily be calculated consistently and faithfully is OPS (On-base Plus Slugging) which is, easily enough, On Base Percentage (OBP) plus Slugging Percentage (SLG). 

But what values are indicative of good performance? For many years casually and conventionally, a .300 hitter was considered good. Someone hitting below .200 was a real liability. A .330 hitter was outstanding. 

While Batting Average and OPS measure very different things, and OPS is more indicative of total offensive contribution, what are the equivalent values for a .200, .300, and .330 hitter?

To answer this I'll evaluate the percentiles of each of those averages over time, and then look to see in the equivalent percentiles of OPS those values fall. 


I can't say for sure how back it goes, but since at least the middle of the 1910s .300 has been described as "a batting mark at which all great players aim"[^1] and "the North Pole of the game"[^2]


## Data and R packages  

```{r}
library(tidyverse)
library(Lahman)
```

## Methodology

Major League Baseball uses the guidleine of 3.1 plate appearances per team game to qualify for league leading batting title for batting average, on-base percentage, and slugging percentage[^3]. From 1920 through 1961 the season was 154 games long (indicating 478 plate appearances required) and from 1962 to the present the season was 162 games long (indicating 503 plate appearances required)[^4]. 


We'll use for [^5,6,7]


```{r}
batter_season_summary <-
  Batting %>%
  filter(yearID >= 1920) %>%
  group_by(yearID,playerID) %>%
  summarise(total_ab = sum(AB),
            total_h = sum(H),
            total_2b = sum(X2B),
            total_3b = sum(X3B),
            total_hr = sum(HR),
            total_bb = sum(BB),
            total_ibb = sum(IBB),
            total_hbp = sum(HBP),
            total_sf = sum(SF),
            total_pa = sum(AB) + sum(BB) + sum(IBB) + sum(SF) + sum(HBP)
            ) %>%
  mutate(avg = total_h/total_ab,
         slg = ((total_h - total_2b - total_3b - total_hr) + 2*total_2b + 3*total_3b + 4*total_hr)/ total_ab,
         obp = (total_h + total_bb + total_ibb + total_hbp)/(total_ab + total_bb + total_ibb + total_hbp + total_sf),
         ops = obp + slg) %>%
  filter((yearID <= 1961 & total_pa >= 478) |
          yearID >= 1962 & total_pa >= 503) 
```


To determine equivalents of a .300 hitter in the other statistical categories we'll need to know where in the distribution of all qualifying batting averages the .300 mark is. And additionally, how has that changed over time?

To accomplish this task computationally use functions from the `purrr` package and the `ecdf()` function from Base R. 


```{r}
three_hundred_hitters <-
batter_season_summary %>%
  split(.$yearID) %>% # from base R
  map_dbl(~ecdf(.$avg)(0.300))

three_thirty_hitters <-
  batter_season_summary %>%
  split(.$yearID) %>% # from base R
  map_dbl(~ecdf(.$avg)(0.330))

two_hundred_hitters <- 
  batter_season_summary %>%
  split(.$yearID) %>% # from base R
  map_dbl(~ecdf(.$avg)(0.200))


length(two_hundred_hitters)
length(three_thirty_hitters)
length(three_hundred_hitters)

cbind(two_hundred_hitters,three_hundred_hitters,three_thirty_hitters)


c(two_hundred_hitters["1974"],three_hundred_hitters["1974"],three_thirty_hitters["1974"])
cbind(two_hundred_hitters,three_hundred_hitters,three_thirty_hitters)["1974",]

result_data <- cbind(two_hundred_hitters,three_hundred_hitters,three_thirty_hitters)
result_data <- as.data.frame(result_data)
result_data <- tibble::rownames_to_column(result_data, "year")
result_data$year <- as.numeric(result_data$year)
head(result_data)
```

```{r}
long_result_data <- 
result_data %>%
  pivot_longer(-year,names_to = "group", values_to = "value") %>%
  filter(year != 1994)
```

# plot!
```{r}
ggplot(data = long_result_data,
       aes(x = year,
           y = value,
           group = group)) +
  geom_line() +
  geom_smooth(method = "lm")
```

```{r}
long_result_data %>%
  group_by(group) %>%
  summarise(mean = mean(value))
```


# now do ops

```{r}
batter_season_summary %>%
  split(.$yearID) %>% # from base R
  map_df(~quantile(x = .$ops, probs = c(.002,.804,.970)))
```



## Any difference between AL and NL


[^1]: St. Nicholas Volume 42. Retrieved January 14, 2020. 
[^2]: The American Magazine, Volume 70. Retrieved January 14, 2010.
[^3]: mlb.com/stats. Retrieved January 14, 2020. 
[^4]: https://en.wikipedia.org/wiki/Major_League_Baseball_schedule#Historical_season_schedules. Retrieved January 14, 2020.
[^5]: https://en.wikipedia.org/wiki/Plate_appearance Retrieved January 14, 2020.
[^6]: https://en.wikipedia.org/wiki/Slugging_percentage Retrieved January 14, 2020.
[^7]: https://en.wikipedia.org/wiki/On-base_percentage Retrieved January 14, 2020.

